---
title: "Project 2: Calibration Plots for COMPAS Recidivism Risk Scores"
author: "Sevastian Sanchez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
    number_sections: yes
    highlight: tango
    theme: default
    fig_caption: yes
    df_print: tibble
  pdf_document:
    toc: no
    toc_depth: '3'
    number_sections: yes
urlcolor: blue
---

# Directory and data loading
```{r include=FALSE}
library(aod)
library(ggplot2)
library(dplyr)

# read data
compas_scores <- read.csv("data/compas-scores-two-years.csv")

# color palette
palette <- c("#46648B", "#F15511", "#D6E1F1", "#000000", "#FFFFFF")
```

# Calibration plot (1) 
predicted vs. observed recidivism with a 45-degree "perfect calibration" reference line
[General crime]
```{r}
risk_scores_vs_recidivism <- compas_scores %>%
  group_by(decile_score.1) %>%
  summarise(
    mean_recid = mean(two_year_recid),
    sd_recid = sd(two_year_recid),
    n = n()
  ) %>%
  mutate(
    observed_rate = mean_recid * 100,
    predicted = decile_score.1 * 10,
    se_rate = (sd_recid / sqrt(n)) * 100
  )

ggplot(risk_scores_vs_recidivism, aes(x = predicted, y = observed_rate)) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", color = palette[4], linewidth = 0.6) +
  geom_errorbar(aes(ymin = observed_rate - se_rate,
                    ymax = observed_rate + se_rate),
                width = 2, color = palette[1], alpha = 0.5) +
  geom_line(color = palette[1], linewidth = 0.9, alpha = 0.7) +
  geom_point(color = palette[1], size = 3.5) +
  scale_x_continuous(breaks = seq(10, 100, 10), limits = c(0, 100)) +
  scale_y_continuous(breaks = seq(0, 100, 10), limits = c(0, 100)) +
  labs(
    title = "General Recidivism Risk Score vs Actual Recidivism",
    x = "Predicted Risk (Score × 10)",
    y = "Observed Recidivism Rate (prop %)",
    caption = "Dashed line = perfect calibration. Error bars = ±1 SE.\nNote: Decile scores rescaled ×10 for visual comparison only; do not represent predicted probabilities."
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    plot.caption = element_text(size = 9, color = "gray50"),
    axis.title = element_text(size = 11),
    panel.grid.minor = element_blank()
  )

ggsave("figures/general_crime_calibration_plot.png", width = 6, height = 4, dpi = 300)

```


# Calibration plot (2)
predicted vs. observed recidivism with a 45° "perfect calibration" reference line 
[Violent crime]
```{r}
risk_scores_vs_violent_recidivism <- compas_scores %>%
  group_by(v_decile_score) %>%
  summarise(
    mean_recid = mean(two_year_recid),
    sd_recid = sd(two_year_recid),
    n = n()
  ) %>%
  mutate(
    observed_rate = mean_recid * 100,
    predicted = v_decile_score * 10,
    se_rate = (sd_recid / sqrt(n)) * 100
  )

ggplot(risk_scores_vs_violent_recidivism, aes(x = predicted, y = observed_rate)) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", color = palette[4], linewidth = 0.6) +
  geom_errorbar(aes(ymin = observed_rate - se_rate,
                    ymax = observed_rate + se_rate),
                width = 2, color = palette[2], alpha = 0.5) +
  geom_line(color = palette[2], linewidth = 0.9, alpha = 0.7) +
  geom_point(color = palette[2], size = 3.5) +
  scale_x_continuous(breaks = seq(10, 100, 10), limits = c(0, 100)) +
  scale_y_continuous(breaks = seq(0, 100, 10), limits = c(0, 100)) +
  labs(
    title = "Violent Recidivism Risk Score vs Actual Recidivism",
    x = "Predicted Risk (Score × 10)",
    y = "Observed Recidivism Rate (prop %)",
    caption = "Dashed line = perfect calibration. Error bars = ±1 SE.\nNote: Decile scores rescaled ×10 for visual comparison only; do not represent predicted probabilities."
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    plot.caption = element_text(size = 9, color = "gray50"),
    axis.title = element_text(size = 11),
    panel.grid.minor = element_blank()
  )

ggsave("figures/violent_crime_calibration_plot.png", width = 6, height = 4, dpi = 300)
```

# Calibrtion Plot (3) -- COMPARASON BY RACE
predicted vs. observed recidivism between black and white individuals with a 45° 
"perfect calibration" reference line 
[General crime]
```{r}
# filter to Black and White respondents only, then compute observed rates per group
calibration_by_race <- compas_scores %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, decile_score.1) %>%
  summarise(
    mean_recid = mean(two_year_recid),
    sd_recid = sd(two_year_recid),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    observed_rate = mean_recid * 100,
    predicted = decile_score.1 * 10,
    se_rate = (sd_recid / sqrt(n)) * 100
  )

ggplot(calibration_by_race, aes(x = predicted, y = observed_rate,
                                 color = race, group = race)) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", color = palette[4], linewidth = 0.6) +
  geom_errorbar(aes(ymin = observed_rate - se_rate,
                    ymax = observed_rate + se_rate),
                width = 2, alpha = 0.4) +
  geom_line(linewidth = 0.9, alpha = 0.8) +
  geom_point(size = 3.5) +
  scale_color_manual(
    values = c("African-American" = palette[1],   # slate blue
               "Caucasian" = palette[2]),   # orange
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(10, 100, 10), limits = c(0, 100)) +
  scale_y_continuous(breaks = seq(0, 100, 10), limits = c(0, 100)) +
  labs(
    title = "Calibration by Race: Predicted vs Observed Recidivism",
    x = "Predicted Risk (Score × 10)",
    y = "Observed Recidivism Rate (prop %)",
    caption = "Dashed line = perfect calibration. Error bars = ±1 SE.\nNote: Decile scores rescaled ×10 for visual comparison only; do not represent predicted probabilities."
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    plot.caption = element_text(size = 9, color = "gray50"),
    axis.title = element_text(size = 11),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )

ggsave("figures/general_calibration_by_race.png", width = 7, height = 5, dpi = 300)

```

# Calibrtion Plot (4) -- COMPARASON BY RACE
predicted vs. observed recidivism between black and white individuals with a 45° 
"perfect calibration" reference line 
[Violent crime]
```{r}
calibration_by_race_violent <- compas_scores %>%
  filter(race %in% c("African-American", "Caucasian")) %>%
  group_by(race, v_decile_score) %>%
  summarise(
    mean_recid = mean(two_year_recid),
    sd_recid = sd(two_year_recid),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(
    observed_rate = mean_recid * 100,
    predicted = v_decile_score * 10,
    se_rate = (sd_recid / sqrt(n)) * 100
  )

ggplot(calibration_by_race_violent, aes(x = predicted, y = observed_rate,
                                         color = race, group = race)) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", color = palette[4], linewidth = 0.6) +
  geom_errorbar(aes(ymin = observed_rate - se_rate,
                    ymax = observed_rate + se_rate),
                width = 2, alpha = 0.4) +
  geom_line(linewidth = 0.9, alpha = 0.8) +
  geom_point(size = 3.5) +
  scale_color_manual(
    values = c("African-American" = palette[1],
               "Caucasian" = palette[2]),
    name   = NULL
  ) +
  scale_x_continuous(breaks = seq(10, 100, 10), limits = c(0, 100)) +
  scale_y_continuous(breaks = seq(0, 100, 10), limits = c(0, 100)) +
  labs(
    title = "Calibration by Race: Violent Recidivism Risk Score vs Actual Recidivism",
    x = "Predicted Risk (Score × 10)",
    y = "Observed Recidivism Rate (prop %)",
    caption = "Dashed line = perfect calibration. Error bars = ±1 SE.\nNote: Decile scores rescaled ×10 for visual comparison only; do not represent predicted probabilities."
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    plot.caption = element_text(size = 9, color = "gray50"),
    axis.title = element_text(size = 11),
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )

ggsave("figures/violent_calibration_by_race.png", width = 7, height = 5, dpi = 300)

```

